Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/hat-vldb.tex
alan intro

\begin{abstract}
To minimize network latency and remain online during partial failure
and network partitions, many modern distributed data storage systems
eschew transactional functionality, which provides strong semantic
guarantees for groups of multiple operations over multiple data
items. In this work, we consider the problem of providing Highly
Available Transactions (HATs), or transactional guarantees that do not
suffer from unavailability during system partitions or incur high
network latency.  We introduce a new taxonomy of high availability and
analyze a existing ACID and distributed consistency guarantees to
identify which can and cannot be achieved in HAT systems. In doing so,
we analytically and experimentally quantify the benefit of providing
high availability as well as the associated semantic compromises.
\end{abstract}

In the debate over NoSQL systems versus traditional DBMS technologies for distributed data management, one key issue is that many NoSQL systems aim for availability in the face of partitions, and provide no, or restricted, support for transactions that group multiple operations over multiple data items. The authors recently advocated Highly Available Transaction (HAT) systems, where general transactions are provided with ACID-like semantic guarantees such as weak isolation levels, in ways that remain available during partitions and avoid high network latency. Here we introduce a new taxonomy of partition-availability properties. We analyse existing ACID and distributed consistency guarantees, showing the availability that can be achieved for each.  We analytically and experimentally estimate the performance gains and semantic compromises of providing high availability, compared to traditional serializable transactions in an internet-scale setting.

>>>>

\begin{abstract}
To minimize network latency and remain online during partial failure
and network partitions, many modern distributed data storage systems
eschew transactional functionality, which provides strong semantic
guarantees for groups of multiple operations over multiple data
items. In this work, we consider the problem of providing Highly
Available Transactions (HATs), or transactional guarantees that do not
suffer from unavailability during system partitions or incur high
network latency.  We introduce a new taxonomy of highly available
systems and analyze existing ACID and distributed consistency
guarantees to identify which can and cannot be achieved in HAT
systems. We analytically and experimentally quantify the performance
and availability benefits of HAT systems as well as the associated
semantic compromises.
\end{abstract}

### 2013-03-31 writeup/intro.tex
alan intro

In July 2000, the CAP Theorem~\cite{gilbert-cap} was crisply
formulated that a highly available system cannot provide strong
consistency guarantees in the presence of network partitions. This
became a rallying call for a generation of highly scalable but
semantically weak system designs. The NoSQL movement embraced this
mantra, leading to many production quality databases that eschewed
strong consistency guarantees, but could scale-out to many
servers~\cite{bigtable, pnuts, dynamo}. The consistency concept used
in the CAP theorem was proposed by the distributed systems community,
and concerned providing strong recency guarantees of a data item that
is replicated across servers.  However, the NoSQL movement implicitly
took it to imply that transactions are also unavailable in the
presence of network partitions. Over a decade later, we observe a
dearth of highly available systems providing general-purpose
multi-operation \textit{transactions} over multiple data items with
useful semantics.

 Serializability, the gold standard of traditional ACID transactions,
 is indeed not achievable with the requirement of high availability
 the in the presence of partitions~\cite{davidson-survey}. However,
 database systems have a long tradition of providing weaker isolation
 and consistency guarantees~\cite{adya, ansicritique,
   gray-isolation}. Due to their concurrency and overall performance
 benefits, weak isolation models are widely employed, overwhelmingly
 as the default setting in today's traditional ``ACID'' and
 up-and-coming ``NewSQL'' databases and often as the maximum guarantee
 offered (Section~\ref{sec:modernacid}). While this weak isolation
 does not provide serializability for general-purpose transactions, it
 is apparently strong enough to deliver acceptable behavior to many
 application programmers and is substantially stronger than the
 typical semantics of current highly available systems. This raises
 the question: which of these semantics can be provided while always
 remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
is poorly understood. We have a strong understanding of weak isolation
in the single-server context from which it originated~\cite{adya,
  ansicritique, gray-isolation} as well as several techniques for
providing distributed serializability~\cite{bernstein-concurrency,
  spanner, granola, daudjee-session, calvin}.  However, the current
literature does not provide guidance as to the operation of weakly
consistent transactional guarantees in a replicated context. To make
matters worse, the distributed computing literature has produced a
large body of work on weak data consistency models for single
replicated objects. Accordingly, in this work, we consider the problem
of providing understanding semantic guarantees regarding multiple
operations operating on data items that are achievable with high
availability and low latency, albeit with some cost to achievable
semantics. While our prior, preliminary results indicated that some
models are achievable~\cite{hat-hotos}, the literature lacks a deep
understanding of what models are available, which are not, and when
this distinction matters.

This work attempts to bridge the gap between, one the one hand, the
plethora of isolation and consistency models proposed in the
transactions and distributed systems literature, and, on the other hand
highly available systems. We introduce a hierarchy of consistency
models and rigorously taxonomize their availability characteristics;
in doing so, we unify previous work on ACID properties~\cite{adya},
session guarantees~\cite{sessionguarantees}, and traditional
distributed ``register'' semantics~\cite{herlihy-art}.

Our main contributions are the following. We classify among the wide
array of isolation and consistency models those that are highly
available, denoting them as {\em Highly Available Transactions
  (HATs)}.  In doing so, we show that many implementations that
provide HAT-semantics are not available, but that the unavailability
is {\em not} inherent to their semantic guarantees. Thus, we provide
highly available implementations for HATs.  Furthermore, our
investigation shows, somewhat surprisingly, that besides
serializability, only Snapshot Isolation (SI) and Repeatable Reads
(RR) are unavailable. The reason for their unavailability stems from
their requirement to avoid update conflict (preventing Lost Updates
and Write Skew phenomena). Finally, we show that many of the weaker
data consistency models from distributed systems can be implemented in
a highly available manner as well. For that, the model of the system
has to be changed to include clients, which cannot be mobile, but
rather have to {\em stick} to one server, something which is implicit
in much of the previous work on distributed systems.

This taxonomy allows us to weigh the costs and benefits of a wide
range of semantic guarantees, but which guarantees are worthwhile in
practice? We attempt to objectively study the virtues and limitations
of both classic and HAT systems by surveying practitioner accounts and
research literature, performing experimental analyses on modern cloud
infrastructure, and analyzing representative applications for their
semantic requirements. While our results are not definitive for all
applications, they suggest that HATs offer a one to three order of
magnitude latency decrease compared to traditional protocols and can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring more
coordination-intensive, unavailable mechanisms. Concisely, we attempt
to understand \textit{what benefits do Highly Available Transactions
  offer and what is their cost?}

>>>>

Brewer's 2000 CAP Theorem stated that a highly available system cannot
provide strong consistency guarantees in the presence of network
partitions~\cite{gilbert-cap}. This result became a rallying call for
a generation of highly scalable but semantically weak system
designs. Most notably, the NoSQL movement embraced this approach,
offering production-use data platforms that eschewed strong
consistency guarantees but offered graceful scale-out to many
servers~\cite{bigtable, pnuts, dynamo}. The consistency concept
employed in the CAP Theorem came from the distributed systems
community and concerned reading the most recent version of a data item
that is replicated across servers.  However, the NoSQL movement
implicitly used this fairly narrow result to justify a lack of
general-purpose transactional functionality, or useful semantics for
groups of operations over multiple data items~\cite{hn,
  foundation-article}. Over a decade after CAP, we observe a dearth of
highly available systems providing transactional functionality.

Serializability, the gold standard of traditional ACID transactions,
is indeed not achievable given a requirement for high availability in
the presence of partitions~\cite{davidson-survey}. However, database
systems have a long tradition of providing weaker isolation and
consistency guarantees~\cite{adya, ansicritique, gray-isolation}. Due
to their concurrency and overall performance benefits, weak isolation
models are widely employed, overwhelmingly as the default setting in
today's traditional ``ACID'' and up-and-coming ``NewSQL'' databases
and often as the maximum guarantee offered
(Section~\ref{sec:modernacid}). While weak isolation levels do not
provide serializability for general-purpose transactions, they are
apparently strong enough to deliver acceptable behavior to many
application programmers and are substantially stronger than the
semantics of current highly available systems. This raises a natural
question: which of these alaternative semantics can be provided while
remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
has been poorly understood. We have a strong understanding of weak
isolation in the single-server context from which it
originated~\cite{adya, ansicritique, gray-isolation} and many papers
offer techniques for providing distributed
serializability~\cite{bernstein-concurrency, spanner, granola,
  daudjee-session, krikellas-bargain, calvin} or snapshot
isolation~\cite{daudjee-snapshot,generalizedsnapshot, kemme-snapshot,
  walter}. The distributed computing, programming language, and
parallel hardware literature also contains many consistency models for
single operations on replicated objects~\cite{pnuts, herlihy-art,
  cops, eiger, cac, sessionguarantees}. However, the literature has
largely ignored semantic guarantees for multiple operations operating
on multiple data items in a highly available distributed environment.

The contributions of this paper fill the gap between, on one hand, the
plethora of previously proposed isolation and consistency models, and,
on the other hand, the goal of high availability. We introduce a
hierarchy of consistency models and, in doing so, unify previous work
on ACID properties~\cite{adya}, session
guarantees~\cite{sessionguarantees}, and traditional distributed
``register'' semantics~\cite{herlihy-art}. We rigorously taxonomize
the availability characteristics of these consistency models: we
determine which can be offered as Highly Available Transactions
(HATs). In doing so, we demonstrate that many \textit{implementations}
that provide these semantics are not highly available but that their
unavailability is not \textit{inherently} required.

Our investigation shows, somewhat surprisingly, that besides
serializability, only fairly strong models like Snapshot Isolation
(SI) and Repeatable Reads (RR) are not highly available, stemming from
their requirement to detect conflicts between concurrent updates (as
needed for preventing Lost Updates or Write Skew phenomena). In
contrast, Read Committed isolation, Transactional Atomicity, and many
of the weaker data consistency models from database and distributed
systems can be implemented in a highly available manner. For several
of these guarantees, such as Transactional Causal Predicate Cut
Isolation, we consider a modified form of high availiabity in which
clients ``stick to'' (i.e. have affinity with) at least one server (a
property which is implicit in the distributed systems literature but
which requires \textit{explicit} consideration in a client-server
replicated database context).

This taxonomy allows us to determine that a wide range of semantic
guarantees can be delivered with availability in face of partitions,
but which guarantees are worthwhile in practice? We attempt to
objectively study the virtues and limitations of both classic and HAT
systems by surveying practitioner accounts and research literature,
performing experimental analyses on modern cloud infrastructure, and
analyzing representative applications for their semantic
requirements. While our results are not definitive, they suggest that
HATs offer a one to three order of magnitude latency decrease compared
to traditional distributed serializability protocols, and they can
provide acceptable semantics for a wide range of programs, especially
those with monotonic logic and commutative updates~\cite{calm, blooml,
  crdt}. HAT systems can also enforce arbitrary foreign key
constraints on multi-item updates and sometimes provide limited
uniqueness guarantees. However, particularly for programs with
non-monotonic logic, HATs may fall short, requiring
sometimes-unavailable mechanisms that are more
coordination-intensive. Overall, this paper quantifies the benefits
Highly Available Transactions can offer as well as the necessary
semantic restrictions they place on applications.
